### 2016-07-03 store-consistency.tex
Typos fixed.

First, we observe that the semantics of a weakly-consistent data store
can be captured by store-specific consistency constraints, along with
transaction-specific isolation constraints, via the trace invariant
$\I$. In particular, we can split $\I$ into two components: (1).
$\I_s$, the store-specific invariant, and (2). $\I_c$, the
program-specific (or, client-specific) invariant, to capture
consistency and isolation constraints, resp.  $\I$ is now a
conjunction: $\I \,=\, \lambda\E.~\I_s(\E) \wedge \I_c(\E)$ (often
simplified to $\I \,=\, \I_s\wedge\I_c$).  While the program-specific
trace invariant ($\I_c$) remains an invariant regardless of the
specific consistency and visibility features exposed by the underlying
data store, the store specific invariant ($\I_s$) changes from store
to store depending on the consisteny level.  For an eventually
consistent store, $\I_s$ is simply \emph{true}. Stronger stores, such
as those that support strong consistenty, have non-trivial definitions
for $\I_s$.

@@

In practice, relational databases resolve such conflicts by
prioritizing weak isolation (thus, concurrency and performance) over
strong consistency, so the execution traces do not necessarily satisfy
{\sc sc}. In particular, visibility constraints imposed by {\sc sc}
are violated iff they are found to be in conflict with the constraints
imposed by a transaction's isolation level. In the context of
the aforementioned example, $\eta_1$ is not made visible to $\eta_2$
because doing so would violate $\I_c$. However, if $\I_c$ is true,
then the store makes $\eta_1$ visible to $\eta_2$ to honor its
consistency commitment\footnote{The term \emph{recency
    commitment}~\cite{bailishat} is often used in practice
  to capture the best-effort nature of {\sc sc}.}. We generalize this
approach to any $\I_s$ and $\I_c$ by defining a \emph{maximum
  visiblity principle} to determine an acceptable weakening of $\I_s$
in case of a conflict with $\I_c$.  The principle requires the
weakened consistency guarantee ($\I_s'$) of the store to enforce all
visibility relationships imposed by the actual consistency guarantee
($\I_s$), unless enforcing such a relationship violates $\I_c$.
% The formal definition of the principle is relegated
% to the supplementary in the interest of space, but it can be
% understood in the context of an {\sc sc} store, where it weakens the
% store invariant to the following:
Formally:
\begin{definition}
$\I_s' : \E \rightarrow \Prop$ is said to be a maximum visibility
weakening of $\I_s : \E \rightarrow \Prop$ if and only if:
\begin{itemize}
  \item $\I_s'$ is weaker than $\I_s$: 
      $\forall\E.~ \I_s(\E) \Rightarrow \I_s'(\E)$, and
  \item In every trace $\E$ that satisfies $\I_c$, and for every pair
  of effects $\eta_1$ and $\eta_2$ in $\E$, if $\I_s(\E)$ requires
  $\eta_1$ to be visible to $\eta_2$, then so does $\I_s'(\E)$ unless
  extending $\E$ with $\visZ(\eta_1,\eta_2)$ violates
  $\I_c$\footnote{\GK{ToDo: consider other well-formedness conditions
  on trace, such as acyclicity of $\visZ$ and $\soZ$. Are they needed
  (considering that the machine never violates them)? Encode the
  specifications in Z3 and make sure they are consistent.}}:
  \begin{smathpar}
  \begin{array}{l}
  \forall\E,\eta_1,\eta_2.~ \I_c(\E) \Rightarrow (\I_s(\E)
    \Rightarrow \underE{\eta_1 \visar \eta_2}) \Rightarrow \\
    \hspace*{0.5in}(\I_s'(\E) \Rightarrow \underE{\eta_1 \visar
    \eta_2} \disj \neg\I_c(\E\,\cup\,(\emptyset,\{(\eta_1,\eta_2)\})))
  \end{array}
  \end{smathpar}
\end{itemize}
\end{definition}
Applying this principle, we can weaken {\sc sc} to obtain the
following store trace invariant ($\I_s$) for an {\sc sc} store whose
isolation constraints are captured by $\I_c$:
\begin{smathpar}
\begin{array}{lcl}
\I_s(\E) & = & \forall \eta_1,\eta_2.\, \{\eta_1,\eta_2\},
    \subseteq \E.\A \conj \id(\eta_1) <
    \id(\eta_2) \\
    & & \hspace*{0.5in} \Rightarrow 
      \underE{\eta_1 \visar \eta_2} \disj \neg\I_c(\E
    \cup (\emptyset,\{(\eta_1,\eta_2)\}))\\
\end{array}
\end{smathpar}
$\I_s$ requires a trace $\E$ to satisfy the visibility constraints of
    {\sc sc} except in cases where they are in conflict with $\I_c$.
    Instantiating the parameter $\I$ with $\I_s \wedge \I_c$ in
    Fig.~\ref{fig:txnimp} results in an operational semantics that
    describes many kinds of relational database-like data stores,
    including causally-consistent data stores, among
    others.\footnote{Details of this instantiation are relegated to
      the supplementary in the interest of space.}

@@

We now consider the proof of the example in Fig.~\ref{fig:motiv-eg-1}
in greater detail. We assume an {\sc sc} store, such as a relational
database, whose store-specific trace invaraint ($\I_s$) was shown
previously. Both transactions are run at {\sc si} isolation, hence
$\I_c$ is $\lambda\E.~\underE{\C{SI(Wd1)}} \conj
\underE{\C{SI(Wd1)}}$. As usual, $\I$ is $\I_s \conj \I_c$.

@@

Decorated parts of \C{Wd1} relevant to the discussion are shown in
Fig.~\ref{fig:wd1-decorated}.  \C{Wd2} is similar and not
shown\footnote{Supplementary contains fully decorated program.}.
Temporary name \C{v3} has been introduced to capture the result of the
RHS expression \C{C-a1}.  Assertions implicitly refer to the current
execution ($\E$), just as hoare triples implicitly refer to the
current state.  The context for propositions is also the implicit,
i.e., we write $\psi$ instead of $\underE{\psi}$. The proposition \C{k
$\ge$ a1+a2} remains an invariant, hence elided. The precondition
$P_1$ of \C{Wd1} accounts for the possibility of \C{Wd2} committing
before \C{Wd1}, writing \C{k-a2} to \C{C}. Since neither \C{Wd1} nor
\C{Wd2} are committed at the beginning, the precondition $P$ of the
program is extended with $\neg\committed(\C{Wd1}) \wedge
\neg\committed(\C{Wd2})$, from which $P_1$ follows. The postcondition
$Q_1$ of \C{Wd1} asserts different values for \C{C} depending on
whether or not \C{Wd2} has already committed.  It also asserts that
\C{Wd1} has committed. The postcondition $Q_2$ of \C{Wd2} (not shown)
is similar, but with the roles of \C{Wd1} and \C{Wd2} exchanged. The
postcondition $Q$ follows from $Q_1 \wedge Q_2$, hence valid. It
remains to show that \C{Wd1} satisfies its specification (\C{Wd2} must
also satisfy its spec, but the proof is similar).

First, we focus on the sequential aspect of the proof and show that
the assertions that decorate \C{Wd1} are indeed valid. We consider the
triple associated with the assignment to \C{C} for illustration. The
precondition $\phi_4$ asserts different values for \C{C} and \C{v3}
(temporary alias of RHS expression) depending on whether or not
\C{Wd2} is committed. Inside \C{Wd1}, $\committed(\C{Wd2})$ may mean
that either \C{Wd2} happened before \C{Wd1} (hence can be visible to
all of \C{Wd1}), or that it committed concurrently with \C{Wd1} (hence
cannot be visible to all of \C{Wd2}).  Nonetheless, $\phi_4$ only
considers the case when $\C{Wd2} \visar \C{Wd1}$. As we show below,
this is sufficient to prove the postcondition $\phi_5$, which appears
to make a stronger claim about the value of \C{C}. The proof follows
from the RG rule \rulelabel{RG-Asgn} for assignments, which allows us
to easily conclude the following about the execution state after the
assignment:

@@

\noindent The rule also lets us assume that the execution satisfies trace
invariant ($\I$), which asserts {\sc si} for both transactions.  Since
both transactions now write to \C{C}, {\sc si} requires either
$\C{Wd1} \visar \C{Wd2}$, or $\C{Wd2} \visar \C{Wd1}$. Since \C{Wd1}
is not yet committed ($\neg\committed(\C{Wd1})$), {\sc si} (which
subsumes {\sc rc}) prohibits the former possibility, allowing us the
deduce $\C{Wd2} \visar \C{Wd1}$. This lets us derive
${\committed(\C{Wd2})}  \Rightarrow \C{C = k-a2} \wedge \C{C =
k-a2-a1}$, allowing us to prove the postcondition $\phi_5$.

@@

The final aspect of the proof is to show that any interference from
\C{Wd1} is contained in its guarantee relation $G_1$, which is also
the rely relation $R_2$ for \C{Wd2}. Since the implementation of
\C{Wd2} is same as \C{Wd1}, $R_2$ (hence $G_1$) is same as $R_1$ shown
above, but with \C{Wd2} and \C{Wd1} interchanged. Continuing our focus
on the assigment statement, the guarantee proof requires us to show
that if executing the assignment takes the execution from $\E$ to
$\E'$, such that $\E'$ satisfies $\I$, then $(\E,\E') \in G_1$.  The
proof follows from the \rulelabel{RG-Asgn} rule, which establishes
that (a). the assignment statement does not commit \C{Wd2}, and (b).
it adds a write effect $\eta$ such that $\txn(\eta)=\C{Wd1}$. This
finishes the proof for the triple in focus. The proofs for other
triples of \C{Wd1} can also be constructed similarly. The combination
of correctness proofs for \C{Wd1} and \C{Wd2} yeilds a correctness
proof for the program.

>>>>

First, we observe that the semantics of a weakly-consistent data store
can be captured by store-specific consistency constraints, along with
transaction-specific isolation constraints, via the trace invariant
$\I$. In particular, we can split $\I$ into two components: (1).
$\I_s$, the store-specific invariant, and (2). $\I_c$, the
program-specific (or, client-specific) invariant, to capture
consistency and isolation constraints, resp.  $\I$ is now a
conjunction: $\I \,=\, \lambda\E.~\I_s(\E) \wedge \I_c(\E)$ (often
simplified to $\I \,=\, \I_s\wedge\I_c$).  While the program-specific
trace invariant ($\I_c$) remains an invariant regardless of the
specific consistency and visibility features exposed by the underlying
data store, the store specific invariant ($\I_s$) changes from store
to store depending on the consistency level.  For an eventually
consistent store, $\I_s$ is simply \emph{true}. Stronger stores, such
as those that support strong consistent, have non-trivial definitions
for $\I_s$.

@@

In practice, relational databases resolve such conflicts by
prioritizing weak isolation (thus, concurrency and performance) over
strong consistency, so the execution traces do not necessarily satisfy
{\sc sc}. In particular, visibility constraints imposed by {\sc sc}
are violated iff they are found to be in conflict with the constraints
imposed by a transaction's isolation level. In the context of
the aforementioned example, $\eta_1$ is not made visible to $\eta_2$
because doing so would violate $\I_c$. However, if $\I_c$ is true,
then the store makes $\eta_1$ visible to $\eta_2$ to honor its
consistency commitment\footnote{The term \emph{recency
    commitment}~\cite{bailishat} is often used in practice
  to capture the best-effort nature of {\sc sc}.}. We generalize this
approach to any $\I_s$ and $\I_c$ by defining a \emph{maximum
  visibility principle} to determine an acceptable weakening of $\I_s$
in case of a conflict with $\I_c$.  The principle requires the
weakened consistency guarantee ($\I_s'$) of the store to enforce all
visibility relationships imposed by the actual consistency guarantee
($\I_s$), unless enforcing such a relationship violates $\I_c$.
% The formal definition of the principle is relegated
% to the supplementary in the interest of space, but it can be
% understood in the context of an {\sc sc} store, where it weakens the
% store invariant to the following:
Formally:
\begin{definition}
$\I_s' : \E \rightarrow \Prop$ is said to be a maximum visibility
weakening of $\I_s : \E \rightarrow \Prop$ if and only if:
\begin{itemize}
  \item $\I_s'$ is weaker than $\I_s$: 
      $\forall\E.~ \I_s(\E) \Rightarrow \I_s'(\E)$, and
  \item In every trace $\E$ that satisfies $\I_c$, and for every pair
  of effects $\eta_1$ and $\eta_2$ in $\E$, if $\I_s(\E)$ requires
  $\eta_1$ to be visible to $\eta_2$, then so does $\I_s'(\E)$ unless
  extending $\E$ with $\visZ(\eta_1,\eta_2)$ violates
  $\I_c$\footnote{\GK{ToDo: consider other well-formedness conditions
  on trace, such as acyclicity of $\visZ$ and $\soZ$. Are they needed
  (considering that the machine never violates them)? Encode the
  specifications in Z3 and make sure they are consistent.}}:
  \begin{smathpar}
  \begin{array}{l}
  \forall\E,\eta_1,\eta_2.~ \I_c(\E) \Rightarrow (\I_s(\E)
    \Rightarrow \underE{\eta_1 \visar \eta_2}) \Rightarrow \\
    \hspace*{0.5in}(\I_s'(\E) \Rightarrow \underE{\eta_1 \visar
    \eta_2} \disj \neg\I_c(\E\,\cup\,(\emptyset,\{(\eta_1,\eta_2)\})))
  \end{array}
  \end{smathpar}
\end{itemize}
\end{definition}
Applying this principle, we can weaken {\sc sc} to obtain the
following store trace invariant ($\I_s$) for an {\sc sc} store whose
isolation constraints are captured by $\I_c$:
\begin{smathpar}
\begin{array}{lcl}
\I_s(\E) & = & \forall \eta_1,\eta_2.\, \{\eta_1,\eta_2\},
    \subseteq \E.\A \conj \id(\eta_1) <
    \id(\eta_2) \\
    & & \hspace*{0.5in} \Rightarrow 
      \underE{\eta_1 \visar \eta_2} \disj \neg\I_c(\E
    \cup (\emptyset,\{(\eta_1,\eta_2)\}))\\
\end{array}
\end{smathpar}
$\I_s$ requires a trace $\E$ to satisfy the visibility constraints of
    {\sc sc} except in cases where they are in conflict with $\I_c$.
    Instantiating the parameter $\I$ with $\I_s \wedge \I_c$ in
    Fig.~\ref{fig:txnimp} results in an operational semantics that
    describes many kinds of relational database-like data stores,
    including causally-consistent data stores, among
    others.\footnote{Details of this instantiation are relegated to
      the supplementary in the interest of space.}

@@

We now consider the proof of the example in Fig.~\ref{fig:motiv-eg-1}
in greater detail. We assume an {\sc sc} store, such as a relational
database, whose store-specific trace invariant ($\I_s$) was shown
previously. Both transactions are run at {\sc si} isolation, hence
$\I_c$ is $\lambda\E.~\underE{\C{SI(Wd1)}} \conj
\underE{\C{SI(Wd1)}}$. As usual, $\I$ is $\I_s \conj \I_c$.

@@

Decorated parts of \C{Wd1} relevant to the discussion are shown in
Fig.~\ref{fig:wd1-decorated}.  \C{Wd2} is similar and not
shown\footnote{Supplementary contains fully decorated program.}.
Temporary name \C{v3} has been introduced to capture the result of the
RHS expression \C{C-a1}.  Assertions implicitly refer to the current
execution ($\E$), just as Hoare triples implicitly refer to the
current state.  The context for propositions is also the implicit,
i.e., we write $\psi$ instead of $\underE{\psi}$. The proposition \C{k
$\ge$ a1+a2} remains an invariant, hence elided. The precondition
$P_1$ of \C{Wd1} accounts for the possibility of \C{Wd2} committing
before \C{Wd1}, writing \C{k-a2} to \C{C}. Since neither \C{Wd1} nor
\C{Wd2} are committed at the beginning, the precondition $P$ of the
program is extended with $\neg\committed(\C{Wd1}) \wedge
\neg\committed(\C{Wd2})$, from which $P_1$ follows. The post-condition
$Q_1$ of \C{Wd1} asserts different values for \C{C} depending on
whether or not \C{Wd2} has already committed.  It also asserts that
\C{Wd1} has committed. The post-condition $Q_2$ of \C{Wd2} (not shown)
is similar, but with the roles of \C{Wd1} and \C{Wd2} exchanged. The
post-condition $Q$ follows from $Q_1 \wedge Q_2$, hence valid. It
remains to show that \C{Wd1} satisfies its specification (\C{Wd2} must
also satisfy its spec, but the proof is similar).

First, we focus on the sequential aspect of the proof and show that
the assertions that decorate \C{Wd1} are indeed valid. We consider the
triple associated with the assignment to \C{C} for illustration. The
precondition $\phi_4$ asserts different values for \C{C} and \C{v3}
(temporary alias of RHS expression) depending on whether or not
\C{Wd2} is committed. Inside \C{Wd1}, $\committed(\C{Wd2})$ may mean
that either \C{Wd2} happened before \C{Wd1} (hence can be visible to
all of \C{Wd1}), or that it committed concurrently with \C{Wd1} (hence
cannot be visible to all of \C{Wd2}).  Nonetheless, $\phi_4$ only
considers the case when $\C{Wd2} \visar \C{Wd1}$. As we show below,
this is sufficient to prove the post-condition $\phi_5$, which appears
to make a stronger claim about the value of \C{C}. The proof follows
from the RG rule \rulelabel{RG-Asgn} for assignments, which allows us
to easily conclude the following about the execution state after the
assignment:

@@

\noindent The rule also lets us assume that the execution satisfies trace
invariant ($\I$), which asserts {\sc si} for both transactions.  Since
both transactions now write to \C{C}, {\sc si} requires either
$\C{Wd1} \visar \C{Wd2}$, or $\C{Wd2} \visar \C{Wd1}$. Since \C{Wd1}
is not yet committed ($\neg\committed(\C{Wd1})$), {\sc si} (which
subsumes {\sc rc}) prohibits the former possibility, allowing us the
deduce $\C{Wd2} \visar \C{Wd1}$. This lets us derive
${\committed(\C{Wd2})}  \Rightarrow \C{C = k-a2} \wedge \C{C =
k-a2-a1}$, allowing us to prove the post-condition $\phi_5$.

@@

The final aspect of the proof is to show that any interference from
\C{Wd1} is contained in its guarantee relation $G_1$, which is also
the rely relation $R_2$ for \C{Wd2}. Since the implementation of
\C{Wd2} is same as \C{Wd1}, $R_2$ (hence $G_1$) is same as $R_1$ shown
above, but with \C{Wd2} and \C{Wd1} interchanged. Continuing our focus
on the assignment statement, the guarantee proof requires us to show
that if executing the assignment takes the execution from $\E$ to
$\E'$, such that $\E'$ satisfies $\I$, then $(\E,\E') \in G_1$.  The
proof follows from the \rulelabel{RG-Asgn} rule, which establishes
that (a). the assignment statement does not commit \C{Wd2}, and (b).
it adds a write effect $\eta$ such that $\txn(\eta)=\C{Wd1}$. This
finishes the proof for the triple in focus. The proofs for other
triples of \C{Wd1} can also be constructed similarly. The combination
of correctness proofs for \C{Wd1} and \C{Wd2} yields a correctness
proof for the program.
